---
editor_options:
  markdown:
    wrap: 150
title: "HADDOCK *Melanogrammus aeglefinus*"
format:
  html:
    theme: theme/advice.scss
    fontsize: 1.125em
    linestretch: 1.9
    embed-resources: true
lang: en
crossref:
  fig-title: ''
  #fig-labels: ''
execute:
  echo: false
  warning: false
  message: false
---

```{r echo=FALSE, warning=FALSE, message=FALSE}
source("config.R")
source("quarto_setup.R")
targets::tar_load(c(
  data_landings,
  data_prog_input,
  data_prognosis,
  data_tac,
  data_assessment
))
targets::tar_load(c(
  advice_hist,
  stock_dev,
  tac_hist
), store = "_assessment_model")
tac <- tac_hist[tac_hist$assessment_year == assessment_year, "tac"]
tac_last_year <- tac_hist[tac_hist$assessment_year == assessment_year - 1, "tac"]
```

```{r}
#| echo: false
#| output: asis

hr_advice_banner(
  publication_date = publication_date,
  tac = tac,
  tac_last_year = tac_last_year,
  year_end = year_end
)
```

## Advice

MFRI and ICES advise that when the Icelandic management plan is applied, catches in the fishing year `r assessment_year`/`r assessment_year+1` should be no more than `r hr_red_dot_number(tac)` tonnes.

## Stock development

Fishing pressure is above HR~MSY~ and HR~pa~ and below HR~lim~; spawning stock size is above MSY B~trigger~, B~pa~ and B~lim~.

```{r}
#| echo: false
library(patchwork)
ggiraph::girafe(ggobj =
  hr_advice_plot_landings(data_landings, assessment_year = assessment_year) +
  hr_advice_plot_recruitment(data_assessment, assessment_year = assessment_year) +
  hr_advice_plot_fpl(data_assessment, assessment_year = assessment_year, ref_points = ref_points) +
  hr_advice_plot_ssb(data_assessment, assessment_year = assessment_year, ref_points = ref_points),
  options = list(
    #ggiraph::opts_hover_inv(css = "opacity:0.7;"),
    #ggiraph::opts_hover(css = "stroke-width:2;"),
    #ggiraph::opts_sizing(width = .7),
    ggiraph::opts_zoom(max = 5)
  ))
```

::: {style="font-size: 11pt;line-height: normal;"}
**Haddock.** Catch by gear type, recruitment, fishing mortality and harvest rate, reference stock biomass (B~45+ cm~) and spawning stock biomass (SSB). Shaded areas and error bars show 95% confidence intervals. All biomass reference points refer to SSB levels.
:::

## Basis of the assessment and reference points
```{r}
hr_advice_basis_table(basis_table)
```

```{r}
#| render: 'asis'
hr_advice_ref_table(ref_points, ref_points_basis_table)
```



## Prospects

The Icelandic Ministry of Industries has adopted a management plan for Icelandic haddock fisheries (MI, 2025). The TAC is set in the following way according to the plan: 

<div class="small-math">

$$
\mathrm{TAC}_{y/y+1} = 0.35B_{45cm+, y} \; \mathrm{if}\; \mathrm{SSB}_{y} \geq \mathrm{MGT \; B_{trigger}}
$$

$$
\mathrm{TAC}_{y/y+1} = \frac{\mathrm{SSB}_{y}}{\mathrm{MGT \; B_{trigger}}} 0.35B_{45cm+, y} \; \mathrm{if}\; \mathrm{SSB}_{y} < \mathrm{MGT \; B_{trigger}}
$$
</div>

where
MGT B~trigger~ = `r hr_red_dot_number(64400)` tonnes of SSB,
*y* is the assessment year,
TAC~y/y+1~ is the TAC for the fishing year starting 01 September in the assessment year,
and B~45cm+,y~ is the estimated biomass of haddock ≥ 45 cm at the beginning of the assessment year

::: {style="font-size: 11pt;line-height: normal;margin-bottom: -2%;"}
**Haddock.** Assumptions made for the interim year and in the forecast.
:::

```{r}
hr_advice_table_prog_input(data_prog_input, assessment_year)
```

::: {style="font-size: 11pt;line-height: normal;margin-bottom: -2%;"}
**Haddock.** Projection of SSB (tonnes) based on adopted harvest control rule.
:::
```{r}
hr_advice_table_prognosis(data_prognosis, assessment_year)
```


## Quality of the assessment

The assessment method was revised in 2025. At the benchmark the model recruitment age was set to be age 1, while previously it was set as age 2. 

```{r}
#| echo: false

#
ggiraph::girafe(ggobj = hr_advice_plot_retro(data_assessment, ref_points, assessment_year),
                options = list(
                  ggiraph::opts_hover_inv(css = "opacity:0.1;"),
                  ggiraph::opts_hover(css = "stroke-width:2;"),
                  #ggiraph::opts_sizing(width = .7),
                  ggiraph::opts_zoom(max = 5)))
```

::: {style="font-size: 11pt;line-height: normal;"}
**Haddock** Current assessment (red line) compared with previous estimates (`r assessment_year - 4`--`r assessment_year - 1`). The recruitment age was changed in 2025.


:::
{{< pagebreak >}}

## Other information

Landings of haddock in Icelandic waters have consistently exceeded the advised catch in recent years, mainly related to quota transfers between species. This results in a harvest rate higher than HR~pa~  but within the realised harvest rates in the evaluation of the management plan. 

## Advice, tac and catch

::: {style="font-size: 11pt;line-height: normal;margin-bottom: -2%;"}
**Haddock.** Recommended TAC, national TAC, and catches (tonnes).
:::

```{r}
hr_advice_table_tac(data_tac)
```

## References and further reading

ICES. 2025. Workshop on the benchmark assessment and management plan evaluation for Icelandic haddock and saithe (WKICEGAD). ICES Scientific Reports.
THIS NEEDS UPDATING WHEN REPORT IS READY1:10. 107 pp. http://doi.org/10.17895/ices.pub.5091

MFRI Assessment Reports 2025. Haddock. Marine and Freshwater Research Institute, 6. June 2025.

Stofnmatsskýrslur Hafrannsóknastofnunar 2025. Ýsa. Hafrannsóknastofnun, 6. júní 2025.

```{r}
#| echo: false
#| output: asis

cat(hr_git_revision(as_html = TRUE))
```
